\documentclass[usletter]{article}
\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{scribe}

\newcommand{\matrixB}[0]{
    \begin{bmatrix}
      0 & 0 \\
      1 & 1 \\
    \end{bmatrix}
}

\begin{document}


\makeheader{John Bender}
           {October 10, 2014}
           {3}
           {Fooling Set Efficacy and The Rank Bound}

\noindent

Fooling sets as defined and described in lecture 2 are one technique for establishing a lower bound on the complexity of the communication protocol for deterministic functions. Here we examine a few examples that highlight how well and how poorly the technique works and we'll also see that, in many cases, the fooling set technique will not provide a good lower bound. We will conclude by introducing the rank bound technique and more examples.

\section{Fooling Set Examples} \label{sec:fooling-set-eg}

\begin{example}
  Greater Than
  \begin{align*}
    & \mathtt{GT}_n : \{0,1\}^n \times \{0,1\}^n \rightarrow \{0,1\}^n \\
    & \mathtt{GT}_n(x,y) = \left \{
      \begin{array}{ll}
        1  & \mbox{if } x \geq y \\
        0 & \mbox{otherwise}
        \end{array}
      \right.
  \end{align*}
\end{example}

First we look at $\mathtt{GT}_n$ which is $1$ when $x \geq y$ for using the integer representations of $x$ and $y$. We can easily choose the fooling set to be $S = \{ (z, z)\ |\ z \in \{0,1\}^n \}$, the set of pairs of inputs with the same bit vector. To see why this is a fooling set consider the characteristic matrix of $\mathtt{GT}_n$ where the rows and columns are ordered by their integer value (note: the fooling set is bolded).

\begin{equation*}
  M_{\mathtt{GT}_n} =
  \begin{bmatrix}
  \textbf{1} & 0 & \cdots & 0 \\
  1 & \textbf{1} & \cdots & 0 \\
  \vdots & \vdots & \ddots & \vdots \\
  1 & 1 & \cdots & \textbf{1}
  \end{bmatrix}
\end{equation*}

\vspace{0.3cm}

This clearly forms a fooling set since each element maps to the same result, $f(z,z) = 1$, and given two pairs $(z_1, z_1), (z_2, z_2)$ either $(z_1, z_2)$ or $(z_2, z_1)$ is in the top right of the matrix and has the value zero. Further since the size of $S$ is $|X|$ and as we saw previously each element in the fooling set must exist in its own f-monochromatic rectangle then, $D(\mathtt{GT_n}) \geq \lceil{log(|S|)}\rceil = n$. We can do slightly better if we also count 0-rectangles $D(\mathtt{GT_n}) \geq \lceil{log(|S| + 1)}\rceil = n + 1$. We know this is tight since the simplest protocol for $\mathtt{GT_n}$ involves one of the participants passing the entirety of their input $|x| \in X = n$ to the other and then a one bit response.

\begin{example}
  Disjointness
  \begin{align*}
    & \mathtt{DISJ}_n : \mathcal{P}({1,\ldots,n}) \times \mathcal{P}({1,\ldots,n}) \rightarrow \{0,1\}^n \\
    & \mathtt{DISJ}_n(x,y) = \left \{
      \begin{array}{ll}
        1  & \mbox{if } x \cap y = \varnothing \\
        0 & \mbox{otherwise}
        \end{array}
      \right.
  \end{align*}
\end{example}

Next we consider $\mathtt{DISJ}_n$, that is the function producing $1$ when $x,y \in \mathcal{P}({1,\ldots,n})$ such that $x \cap y = \varnothing$. Again we select a straight forward fooling set $S = \{(X, \overline{X})\ |\ X \in \mathcal{P}({1,\ldots,n})\}$ where $\overline{X}$ is the compliment of the set $X$. To see that this is a fooling set consider that for any two members of the fooling set, say $(A, \overline{A})$ and $(B, \overline{B})$, $A$ and $B$ must be different. That is,

\begin{equation*}
  \exists z.(z \in A \land z \notin B) \lor (z \in B \land z \notin A)
\end{equation*}

Either case implies that $z$ is in the compliment of $B$ or $A$ respectively, which in turn implies that a pair at the corner of the corresponding rectangle $(A, \overline{B})$ (resp. $(B, \overline{A})$) is not disjoint and the value of the disjointness function would be 0 as required. We conclude that the complexity is bounded from below by $\lceil{log(2^n + 1)}\rceil = n + 1$ which is again tight due to the trivial implementation of the protocol.

\begin{example}
  Inner Product
  \begin{align*}
    & \mathtt{IP}_n : \{0,1\}^n \times \{0,1\}^n \rightarrow \{0,1\} \\
    & \mathtt{IP}_n(x, y) = \bigoplus_{i=1}^n x_i \land y_i
  \end{align*}
\end{example}

Finally we examine the inner product function, $\mathtt{IP}_n$, where the fooling set method fails to provide a good lower bound. $\mathtt{IP}_n$ takes the bitwise $\land$ of each bit position from two bit vectors of length $n$ and then performs a fold over the resulting bit vector with $\oplus$. The output of the function is the single bit result.

To construct the fooling set for $\mathtt{IP}_n$ we pick one of the $n$ bits, flip it and pair it with its duplicate, i.e. $S = \{ (x', x') | i \in 1 \ldots n, x' = \langle \0_1, \ldots 0_{i-1},1,0_{i+1}, \ldots, 0_n \rangle \}$. Each of these pairs clearly results in the desired value and when paired with an other $x'$ the $\land$ of each bit will result in all zeros due to the unique position of the flipped bit in each vector. Unfortunately, because we can only reliably use a single bit to achiever our ends there are only $n$ pairs which gives us the weaker lower bound $log_2(n+1)$.

\section{Fooling Sets For Random Functions} \label{sec:fooling-set-random}

Since we have a sense that fooling sets might not work particularly well for some functions it might be nice to get a better picture of how frequently they are effective for finding a lower bound. We will consider the probability that a fooling set for a random function meets some basic threshold. In particular we will prove the following theorem.

\begin{theorem}
For a random function $f : \{0,1\}^n \times \{0,1\}^n \rightarrow \{0,1\}$ and the function that takes functions to their fooling sets, $fs : (\{0,1\}^n \times \{0,1\}^n \rightarrow \{0,1\}) \rightarrow S$, we have

\begin{equation*}
\Pr\left[ |fs(s)| \geq 10n \right] \leq \left(\frac{2}{3}\right)^{n^{2}}
\end{equation*}
\end{theorem}

That is, we will show that the probability that the size of the fooling set exceeds $10n$ for most random functions is very low. This is also the probability that a corresponding protocol will have a lower bound of greater than $log_2(10n)$. Coupled with the high probability we derived in the first lecture that the complexity of a given protocol will be greater than or equal to $n$ (the size of the inputs) we will be left with low confidence in the fooling set approach.

\begin{proof}
  First we establish the probability that given $S$ is a fooling set for some function $f$. We start by fixing $S = \{(x_1, y_1), \ldots, (x_s, y_s)\}$ such that all $x$ are distinct and all $y$ are distinct and establishing a probability for whether that $S$ is a fooling set for $f$.

\begin{align*}
\Pr_f\left[  S \in \mathsf{FS}(f) \right] &=
  \left( \frac{2}{2^{|S|}} \right) \cdot \left( \frac{3}{4} \right)^{{|S| \choose 2}}
\end{align*}

Where $\mathsf{FS}(f)$ is the set of all fooling sets for $f$, the probability that $S$ is in that set is determined by the probability of two independent events that align with the definition of a fooling set. First, $\left( \frac{2}{2^{|S|}} \right)$ is the probability that the fooling sets combination of $x$s and $y$s all result in the same output result from $f$, i.e. all zeros or all ones. Second, $\left( \frac{3}{4} \right)^{{|S| \choose 2}}$ is the probability that the other corners of the rectangle formed by any two pairs of inputs from $S$ will contain the ``offending'' value. That is, there are three ways that the other two corners could be configured with at least one offending value (e.g. assuming the value for the pairs in $S$ is 1 then $(0,1)$, $(0,0)$, $(1,0)$) which we raise to the power of the number of ways to choose any two elements in $S$ to form the rectangle because for every pairing of elements in $S$ they must exhibit this property.

\begin{align*}
\Pr_f\left[  S \in \mathsf{FS}(f) \right]
  &= \left( \frac{2}{2^{|S|}} \right) \cdot \left( \frac{3}{4} \right)^{{|S| \choose 2}} \\
  &= 2 \left( \frac{1}{\sqrt{3}} \right)^{|S|} \cdot \left( \frac{3}{2} \right)^{|S|^2} \\
  &\ll \left( \frac{3}{2} \right)^{|S|^2}
\end{align*}

We also know that the number of candidate fooling sets for a function $f$ is ${2^n \choose |S|}^2$ where the two choices are for a set of $x$s and a set of $y$s. As a result we can establish the probability that $f$ has a fooling set $S$ of at least $|S|$.

\begin{align*}
\Pr_f\left[ \exists S. S \in \mathsf{FS}(f) \land |S| > x \right]
  &= \Pr_f\left[ \exists S. S \in \mathsf{FS}(f) \land |S| = x \right] \\
  &= {2^n \choose x}^2 \cdot \left( \frac{3}{2} \right)^{x^2} \\
  &= 2^{2nx - log_2(\frac{2}{\sqrt{3}})x^2} \\
  &< \left( \frac{2}{3} \right)^{n^2} \mathsf{for}\ x = 10n
\end{align*}
\end{proof}

So we are left with very little in the way of confidence with the fooling set if we wish to prove a lower bound for a function we are interested in. Next we'll examine a different technique that relies on establishing the rank of the characteristic matrix for the function under consideration.

\section{Matrix Rank} \label{sec:matrix-rank}

\begin{definition}
  Matrix rank is the size of the largest set of linearly independent rows (or columns) for a given matrix. For $M \in F^{n \times m}$ where $F$ is an arbitrary field $rk_{F}(M) = min \{ k' : M = AB, A \in F^{n \times k}, B \in F^{k \times m} \}$. When the field subscript is omitted the field is assumed to be the reals.
\end{definition}

We are primarily interested in boolean matrices (i.e. matrices over the field $F_2$) and the impact of selecting this (or other) fields will be considered in later lectures. To make this distinction clear consider the following matrix:

\begin{example}
Matrix Rank:
\begin{equation*}
  M =
  \begin{bmatrix}
    1 & 1 & 1 & 1 \\
    1 & 1 & 0 & 0 \\
    1 & 0 & 1 & 0 \\
    1 & 0 & 0 & 1
  \end{bmatrix}
\end{equation*}
\end{example}

Here the rank of the matrix over $F_2$ is 3 since the first row can be constructed as a linear combination of the last three (recall that $1 + 1 + 1 = 0$ in $F_2$), but in the rank of the matrix over the reals is four since none of the rows can be constructed using a combination of the others.

Further consider the identity matrix and the matrix of ones, where the rank values are over any field.

\begin{example}
Identity and Ones Matrices:
\begin{align*}
  I_n =
  \begin{bmatrix}
  1 & 0 & \cdots & 0 \\
  0 & 1 & \cdots & \vdots \\
  \vdots & \vdots & \ddots & 0 \\
  0 & \cdots & 0 & 1
  \end{bmatrix}
  &&
  J_n =
  \begin{bmatrix}
  1 & 1 & \cdots & 1 \\
  1 & 1 & \cdots & \vdots \\
  \vdots & \vdots & \ddots & 1 \\
  1 & \cdots & 1 & 1
  \end{bmatrix}
  \\
  \\
  rk_{F}(I_n) = n
  &&
  rk_{F}(J_n) = 1
\end{align*}
\end{example}

Next we define two operations of interest in later discussions and how the rank of matrices changes (or doesn't) under those operations.

\begin{definition}
  Tensor Product:
  \begin{equation*}
    A \otimes B = [A_{i,j}B_{i',j'}]_{(i, i'),(j,j')}
  \end{equation*}
\end{definition}

\begin{example}
  Tensor Product Example:
  \begin{equation*}
    \begin{bmatrix}
      0 & 1 \\
      1 & 0 \\
    \end{bmatrix}
    &&
    \otimes
    \matrixB
    &&
    =
    \begin{bmatrix}
      0 \cdot \matrixB & 1 \cdot \matrixB \\
      1 \cdot \matrixB & 0 \cdot \matrixB
    \end{bmatrix}
  \end{equation*}
\end{example}

\begin{definition}
  Entry-wise Product:

  \begin{equation*}
    A \cdot B = [A_{i,j}B_{i,j}]
  \end{equation*}
\end{definition}

\begin{example}
  Entry-wise Product Example:
  \begin{equation*}
    \begin{bmatrix}
      0 & 1 \\
      1 & 0 \\
    \end{bmatrix}
    &&
    \cdot
    \matrixB
    &&
    =
    \begin{bmatrix}
      0 \cdot 0 & 1 \cdot 0 \\
      1 \cdot 1 & 0 \cdot 1
    \end{bmatrix}
  \end{equation*}
\end{example}

For each of these operations and for matrices over fields we have the following facts:

\begin{fact}

\end{fact}


\bibliographystyle{abbrv}
\bibliography{1}

\end{document}
